### 2010-03-05 article/Article.tex
first words for the article
Ignore-this: b6a750ffac9cb51b7f801b784674e0bb

darcs-hash:20100305021824-2fc9d-535a42c015a545992c83df5464ea80287a8f939a.gz

%
% Complete documentation on the extended LaTeX markup used for Insight
% documentation is available in ``Documenting Insight'', which is part
% of the standard documentation for Insight.  It may be found online
% at:
%
%     http://www.itk.org/

\documentclass{InsightArticle}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}
% to be able to use options in graphics
\usepackage{graphicx}
% for pseudo code
\usepackage{listings}
% subfigures
\usepackage{subfigure}

%  This is a template for Papers to the Insight Journal. 
%  It is comparable to a technical report format.

% The title should be descriptive enough for people to be able to find
% the relevant document. 
\title{FFT based convolution}

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
% \release{0.00}

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Ga\"etan Lehmann{$^1$}}
\authoraddress{{$^1$}INRA, UMR 1198; ENVA; CNRS, FRE 2857, Biologie du
D\'eveloppement et 
Reproduction, Jouy en Josas, F-78350, France.}

\begin{document}
\maketitle

\ifhtml
\chapter*{Front Matter\label{front}}
\fi

\begin{abstract}
\noindent

The Fourier transform of the convolution of two images is equal to the product of their
Fourier transform. With this definition, it is possible to create a convolution filter
based on the Fast Fourier Transform (FFT). The interesting complexity characteristics of
this transform gives a very efficient convolution filter for large kernel images.

This paper provides such a filter, as well as a detailed description of the implementation
choices and a comparison with the "simple" itk::ConvolutionImageFilter.

\end{abstract}

\tableofcontents

\section{Introduction}

The convolution of two image is a very cpu demanding task, with a complexity of $O(N*M)$ where $N$ is
the number of pixel of the first image and $M$ the number of pixels of the second. Usually, the size
of one of the image is a lot smaller than the one of the other -- the smaller is often called the
kernel. The quite small size of the kernel makes the convolution computable in a very usable time,
but when the size of the kernel grows, the computation time quickly becomes impraticable.

Fortunately, the convolution of two images is simply the product of those two images pixel wise in
the frequency domain, with a complexity of $O(max(N, M))$. The size of the kernel, which is usually
way smaller than the image, has no effect on the computation time. The cost of the Fourier Transform
can be quite high however -- that's why the FFT based convolution is more efficient only for big
kernel images.

\section{FFT based convolution step by step}

The FFT based convolution requires several step to be performed. They are detailed one by one.

\subsection{Zero padding}

The images will be padded with zeros for several reasons:
\begin{itemize}
  \item to make their size match. This is required to perform the multiplication in the frequency
  domain.
  \item  to avoid the border effects. Because the FFT considere the image as a cyclic signal, the
  image must be padded to avoid the border effects. The final image size for all the dimensions
  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $P$ the size
  of the other image on the same dimension.
  \item to make the FFT possible, or to enhance its performance. Some FFT implementations, like the
  VNL one, can only be run on an image where the size on all the dimensions are a power of two.
  Some other implementations, like FFTW's one, are performing very differently depending on the
  size of the image. Adding a few more pixels on the border of the image can lead them to perform
  a lot better.
\end{itemize}

\subsection{Normalization}

To preserve the intensity of the pixel in the convolved image, the kernel must be normalized to one
-- the sum of all its pixel is one.

\subsection{Flipping}

The convolution is a flipping transform by nature: the shape of the kernel image can be found in the
convolved image, but flipped on all the axes. To get the same behavior when computing the product
in the frequency domain, the kernel image must be flipped prior to the FFT.

\subsection{Shifting (centering)}

The convolution is usually done with a centered kernel. If the kernel is not centered, the input image
is shifted in the convolved image, by the same shift than the shift of the kernel. However, for the
FFT, the center pixel is the one in the corner, not the one in the center of the image, as is it 
usual for a human, and so the kernel must be shifted.
This shift must be done after the padding, and before the FFT.

Another option, not used here, is to shift the image after the inverse Fourier transform.

\subsection{Fourier transform}

The Fourier transform is performed on the padded input image, and on the normalized, padded, flipped
and shifted kernel.

\subsection{Frequency domain multiplication}

The convolution is performed in the frequency domain, simply by computing the product of the FFT of
the image and of the FFT on the kernel pixel wise.

\subsection{Inverse Fourier transform}

The convolved image in the frequency domain is transformed to the space domain with an inverse Fourier
transform.

\subsection{Cropping}

The image is cropped to fit the size of the input image.

\section{Implementation}

The step by step transform described in the previous chapter can perfectly be implemented in a pure
ITK pipeline model. The few filters which were missing for this task has been implemented, and some
other has been enhanced to improve their performance.

\subsection{NormalizeToConstantImageFilter}

The normalization to one cannot be done without going outside of the ITK pipeline model at this time.
itk::NormalizeToConstantImageFilter has been added for this task.

This filter is implemented as a minipipeline of two filters:
\begin{itemize}
  \item a itk::StatisticsImageFilter, to compute the sum of the pixels in the image;
  \item a itk::DivideByConstantImageFilter, to actually normalize the pixel values.
\end{itemize}
The use of itk::StatisticsImageFilter only to compute the sum might be a littel overkill. This filter
is only used on the kernel which is usually quite small, so the performance impact on the whole
FFT based convolution is small.

\subsection{FFTZeroPaddingImageFilter}

The padding step, as described in the previous chapter, has several goals, which leads to a single
image size used to pad both the input image and the kernel.
All the logic is implemented in itk::FFTZeroPaddingImageFilter.

This filter is in charge of padding both images. It requires two input -- the image to convolve and
the kernel. The pad size is computed according to both image sizes.

The image to convolve is simply padded with zeros. The kernel requires a little more work to be
centered in the padded image and to make its region match the one of the padded image to convolve.
The padding and region changes are implemented as a minipipeline of itk::ConstantPadImageFilter and
itk::ChangeInformationImageFilter.

itk::FFTZeroPaddingImageFilter is also able to extend the padded region in order to enhance the FFT
performance, or to make it possible when the FFT possible when it require a size which is a power of
two. FFTW is able to work on any image size, but due to the algorithms useds, produce significantly
more performant result with some specific sizes. The two main reasons are:
\begin{itemize}
  \item the decomposition composite sizes in smaller transforms using the Cooley–Tukey algorithm;
  \itme the hard coded loop unrolling for the size which are prime numbers up to thirteen.
\end{itemize}
For these reasons, FFTW performs better when the size on all the dimension has its greatest prime
factor smaller or equal to thirteen.
The greatest prime factor can be simply computed with the algorithm described in figure \ref{greatestPrimeFactor}.

\begin{pseudocode}{isPrime}{n}
\FOREACH x \in [2, \sqrt{n}] \DO 
\BEGIN
  \IF n%x \eq 0
  \THEN
  \BEGIN
    \RETURN \FALSE
  \END 
\END
\RETURN \TRUE
\end{pseudocode}

\begin{pseudocode}{greatestPrimeFactor}{n}
v \GETS 2
\WHILE x \leq n \DO
\BEGIN
  \IF n%x \eq 0 \AND \CALL{isPrime}{v}
  \THEN
  \BEGIN
    \IF n \eq v
    \THEN
      \RETURN \FALSE
    \END
    n \GETS n / v
    v \GETS 2
  \ELSE
    v \GETS v + 1
  \END 
\END
\end{pseudocode}

\subsection{FFTW filters enhancements}
\subsubsection{Thread support}
fft et norm
\subsubsection{Memory usage reduction}
\subsubsection{Code safety}

\subsection{RegionFromReferenceImageFilter}

\subsection{FFTConvolutionImageFilter}

\section{Performance}

\subsection{Kernel size}

\subsection{Greatest prime factor}

\subsection{Number of threads}

% \url{http://www.itk.org}
% \code{Insight/Documentation/Style.pdf}
% \section{Principles of Solar Spot Detection}
% \cite{ITKSoftwareGuide}.
% \doxygen{ImageToImageFilter}

% \small \begin{verbatim}
% \end{verbatim} \normalsize

% The {itemize} environment uses a bullet for each \item.  If you want the 
% \item's numbered, use the {enumerate} environment instead.
% \begin{itemize}
%   \item  Insight Toolkit 2.4.
%   \item  CMake 2.2
% \end{itemize}

% \ref{cthead1}

\begin{figure}[htbp]
\centering
\includegraphics{cthead1}
\caption{The input image.\label{cthead1}}
\end{figure}

\appendix

\bibliographystyle{plain}
\bibliography{InsightJournal}
\nocite{ITKSoftwareGuide}

\end{document}

### 2010-03-06 itkFFTZeroPaddingImageFilter.h
enhance greatestPrimeFactor implementation
Ignore-this: 1e2b952470fbcc646f49122fc954e2ff

darcs-hash:20100306022750-2fc9d-e8bd900c815ee5be152308f1784df67684521325.gz

@@ -203,20 +203,14 @@ private:
       {
       if( n%v == 0 && isPrime( v ) )
         {
-        if( n == v )
-          {
-          return v;
-          }
         n /= v;
-        v = 2;
         }
       else
         {
         v += 1;
         }
       }
-    // shouldn't be defined, but it's easier here to simply return the value
-    return n;
+    return v;
     }
 
 }; // end of class

### 2010-03-06 article/Article.tex
article continued
Ignore-this: 7c5a69dcaa860df984460bc2d28eeadd

darcs-hash:20100306024350-2fc9d-1f6ff65f236e08c37a5df96786807a7adbdc4fca.gz

@@
 \usepackage{subfigure}
+\usepackage{pseudocode}
 
@@
 
-The convolution of two image is a very cpu demanding task, with a complexity of $O(N*M)$ where $N$ is
+The convolution of two image is a very cpu demanding task, with a complexity of $O(N \times M)$ where $N$ is
 the number of pixel of the first image and $M$ the number of pixels of the second. Usually, the size
@@
 the frequency domain, with a complexity of $O(max(N, M))$. The size of the kernel, which is usually
-way smaller than the image, has no effect on the computation time. The cost of the Fourier Transform
+way smaller than the image, has no effect on the computation time, so the complexity is simply $O(N)$.
+The cost of the Fourier Transform
 can be quite high however -- that's why the FFT based convolution is more efficient only for big
@@
   image must be padded to avoid the border effects. The final image size for all the dimensions
-  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $P$ the size
+  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $Q$ the size
   of the other image on the same dimension.
@@
   size of the image. Adding a few more pixels on the border of the image can lead them to perform
-  a lot better.
+  a lot better, as shown in chapter \ref{perf:gpf}.
 \end{itemize}
@@
 FFT, the center pixel is the one in the corner, not the one in the center of the image, as is it 
-usual for a human, and so the kernel must be shifted.
+usual for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
+of the image as expeced by the FFT.
+
 This shift must be done after the padding, and before the FFT.
@@
 The step by step transform described in the previous chapter can perfectly be implemented in a pure
-ITK pipeline model. The few filters which were missing for this task has been implemented, and some
-other has been enhanced to improve their performance.
+ITK pipeline model. The few filters which were missing for this task have been implemented, and some
+other have been enhanced to improve their performance.
 
@@
 \end{itemize}
-The use of itk::StatisticsImageFilter only to compute the sum might be a littel overkill. This filter
-is only used on the kernel which is usually quite small, so the performance impact on the whole
+The use of itk::StatisticsImageFilter only to compute the sum might be {\em a little overkill}. This filter
+is only used on the kernel, which is usually quite small, so the performance impact on the whole
 FFT based convolution is small.
@@
 
-The padding step, as described in the previous chapter, has several goals, which leads to a single
+The padding step, as described in the previous chapter, has several goals, which lead to a single
 image size used to pad both the input image and the kernel.
@@
 
-The image to convolve is simply padded with zeros. The kernel requires a little more work to be
-centered in the padded image and to make its region match the one of the padded image to convolve.
-The padding and region changes are implemented as a minipipeline of itk::ConstantPadImageFilter and
+The image to convolve is simply padded with zeros. The kernel, however, requires a little more work to be
+centered in the padded image, and to make its region match the one of the padded image to convolve.
+The padding and the region changes are implemented as a minipipeline of itk::ConstantPadImageFilter and
 itk::ChangeInformationImageFilter.
@@
 \begin{itemize}
-  \item the decomposition composite sizes in smaller transforms using the Cooley–Tukey algorithm;
-  \itme the hard coded loop unrolling for the size which are prime numbers up to thirteen.
+  \item the decomposition of images of composite sizes in smaller transforms using the
+  Cooley-Tukey algorithm;
+  \item the hard coded loop unrolling for the size which are prime numbers up to thirteen.
 \end{itemize}
-For these reasons, FFTW performs better when the size on all the dimension has its greatest prime
+For these reasons, FFTW performs better when the size on each dimension has its greatest prime
 factor smaller or equal to thirteen.
-The greatest prime factor can be simply computed with the algorithm described in figure \ref{greatestPrimeFactor}.
+The greatest prime factor can be simply computed with the algorithm described in algorithm \ref{greatestPrimeFactor}.
 
@@
 \BEGIN
-  \IF n%x \eq 0
-  \THEN
-  \BEGIN
-    \RETURN \FALSE
-  \END 
-\END
+  \IF n \% x = 0 \THEN \RETURN \FALSE \\
+\END \\
 \RETURN \TRUE
@@
 \begin{pseudocode}{greatestPrimeFactor}{n}
-v \GETS 2
+\label{greatestPrimeFactor}
+x \GETS 2 \\
 \WHILE x \leq n \DO
 \BEGIN
-  \IF n%x \eq 0 \AND \CALL{isPrime}{v}
+  \IF n \bmod x = 0 \AND \CALL{isPrime}{x}
   \THEN
-  \BEGIN
-    \IF n \eq v
-    \THEN
-      \RETURN \FALSE
-    \END
-    n \GETS n / v
-    v \GETS 2
+    n \GETS n / x
   \ELSE
     v \GETS v + 1
-  \END 
-\END
+\END \\
+\RETURN x
 \end{pseudocode}
@@
 \subsection{Greatest prime factor}
+\label{perf:gpf}
 

### 2010-03-06 article/Article.tex
article continuated
Ignore-this: f0d0b161df09c007b5e4da7133fcf2a1

darcs-hash:20100306233314-2fc9d-b3a8c2d79931cf5617e6916014410ccc2e80947f.gz

@@
 \subsection{FFTW filters enhancements}
+
+ITK provides FFT and inverse FFT filters based on the FFTW library. Those filters had some problems,
+or were missing some features. They have been enhanced to fit the needs of the FFT based convolution
+filter.
+
 \subsubsection{Thread support}
-fft et norm
+The FFTW filters were not multithreaded. There are two places were multithreading can take place: the
+FFTW execution and the normalization step.
+
+The FFTW execution can be threaded simply by calling $fftwf_plan_with_nthreads(threads);$ or $fftwf_plan_with_nthreads(threads);$, depending on the pixel type, before creating the plan. 
+itk::fftw::Proxy has been modified to provide this feature: an extra $thread$ parameter, which default
+to one to keep the backward compatibility, has been added to the method in charge of creating the plans.
+Then the itk::FFTWComplexConjugateToRealImageFilter and itk::FFTWRealToComplexConjugateImageFilter have
+been modified to pass their user defined number of threads to those methods.
+
+In order to implement the threading support in the normalization step in 
+itk::FFTWComplexConjugateToRealImageFilter, the execution of the FFTW library has been moved to 
+BeforeThreadedGenerateData(), and the normalization step to ThreadedGenerateData().
+
 \subsubsection{Memory usage reduction}
+
+FFTW filters were using a lot of memory, because they were allocating an internal buffer for the input 
+image, and an other one for the output image. Worth, these buffer were not deallocated at the end of
+the execution filter, keeping the memory usage very high even when the execution has completed.
+
+I guess the main reasons to do that were:
+\begin{itemize}
+  \item make sure that the buffers are properly aligned.
+  \item avoid the destruction of the input. The creation of the FFTW plan, and the execution of FFTW 
+  can destroy the input. This is most of the time not acceptable in ITK.
+  \item reuse the FFTW for a latter run. The plan include the memory location of the input and output
+  buffer, so the buffer can't be deallocated, as it may be the case with the filter input and output images.
+\end{itemize}
+
+The buffer were simply allocated with the $new$ operator, which is not enough to ensure a proper alignement
+on some plateforms, including Microsoft Windows and Linux.
+
+The destruction of the input can be avoided in the creation plan by using the $FFTW_ESTIMATE$ flag
+-- that flag was already used in the implementation. It can also be avoided while running FFTW, for 
+the real to complex transform, by also using the flag $FFTW_PRESERVE_INPUT$, so in 
+itk::FFTWRealToComplexConjugateImageFilter, there is no need for the input buffer. The case is a bit
+more difficult for the complex to real transform: FFTW does not provide any algorithm able to preserve
+the input, so the input must be copied to an internal buffer. There is a case however, when we don't
+care about destroying the input in ITK: when the ReleaseDataFlag is ON on the input. In that case, 
+itk::FFTWRealToComplexConjugateImageFilter has been modified to use the input image directly without 
+intermediate buffer. The detection of the ReleaseDataFlag is a bit difficult during the pipeline 
+execution: this flag is always modified to OFF before the call to $GenerateData()$ to avoid the 
+destruction by a minipipeline. In consequence, the detection of the ReleaseDataFlag is done at an 
+earlier stage in the pipeline execution, in $UpdateOutputData()$, and stored in a member variable.
+
+The latter reuse of the plan, while sensible when not using the $FFTW_ESTIMATE$ flag, is not much 
+interesting when it is used: the time needed to create the plan is very short -- way shorter than the
+FFT computation. Also, the number of threads and the image size are stored in the plan, and can't be 
+changed later, which doesn't fit well with ITK pipeline model. In the enhanced version, the plan is
+not kept between the filters execution, and so no intermediate output buffer is used, decreasing 
+significantly the needed memory.
+
 \subsubsection{Code safety}
+
+Only $fftwf_execute(plan);$ and $fftw_execute(plan);$ are thread safe in FFTW -- the other functions
+are not thread safe. In consequence, the call to all the other functions has been protected by a 
+global lock in the enhanced classes.
+
+Also, the check for the size of the image to know if the plan can be reused may not have been good 
+enough, because it was testing only the number of pixels and not the size on all the dimensions. 
+The plan is not kept between the execution, so this can be a problem anymore.
 

### 2010-03-07 article/Article.tex
article continued
Ignore-this: 26996452fdf987ebfa8b4427885f5323

darcs-hash:20100307220037-2fc9d-cdc8c19d63e2e0382b190d594ee38cb974fcd5ba.gz

--- 2010-03-06 article/Article.tex
@@
 \subsection{FFTW filters enhancements}
+
+ITK provides FFT and inverse FFT filters based on the FFTW library. Those filters had some problems,
+or were missing some features. They have been enhanced to fit the needs of the FFT based convolution.
+
 \subsubsection{Thread support}
-fft et norm
+The FFTW filters were not multithreaded. There are two places where multithreading can take place: the
+FFTW execution and the normalization step.
+
+The FFTW execution can be threaded simply by calling \verb$fftwf_plan_with_nthreads(threads);$ or 
+\verb$fftwf_plan_with_nthreads(threads);$, depending on the pixel type, before creating the plan. 
+\verb$itk::fftw::Proxy$ has been modified to provide this feature: an extra \verb$thread$ parameter,
+which defaults
+to $1$ to keep the backward compatibility, has been added to the method in charge of creating the plans.
+Then the \verb$itk::FFTWComplexConjugateToRealImageFilter$ and
+\verb$itk::FFTWRealToComplexConjugateImageFilter$ have
+been modified to pass their user defined number of threads to those methods.
+
+In order to implement the threading support in the normalization step in 
+\verb$itk::FFTWComplexConjugateToRealImageFilter$, the execution of the FFTW library has been moved to 
+\verb$BeforeThreadedGenerateData()$, and the normalization step to \verb$ThreadedGenerateData()$.
+
 \subsubsection{Memory usage reduction}
+
+FFTW filters were using a lot of memory, because they were allocating an internal buffer for the input 
+image, and an other one for the output image. Worth, these buffers were not deallocated at the end of
+the execution filter, keeping the memory usage very high even when the execution has completed.
+
+I guess the main reasons to do that were:
+\begin{itemize}
+  \item avoid the destruction of the input. The creation of the FFTW plan, and the execution of FFTW 
+  can destroy the input. This is most of the time not acceptable in ITK.
+  \item reuse the FFTW plan for a latter run. The plan include the memory location of the input and output
+  buffers, so the buffers can't be deallocated, as it may be the case with the filter's input and output images.
+\end{itemize}
+
+The destruction of the input can be avoided in the creation plan by using the \verb$FFTW_ESTIMATE$ flag
+-- that flag was already used in the implementation. It can also be avoided while running FFTW, for 
+the real to complex transform, by also using the flag \verb$FFTW_PRESERVE_INPUT$. So in 
+\verb$itk::FFTWRealToComplexConjugateImageFilter$, if the right flags are used, there is no need for
+the input buffer.
+The case is a bit
+more difficult for the complex to real transform: FFTW does not provide any algorithm able to preserve
+the input, so the input must be copied to an internal buffer. There is a case however, when we don't
+care about destroying the input in ITK: when the \verb$ReleaseDataFlag$ is \verb$ON$ on the input of the filter.
+In that case, the input will be deallocated right after the end of the execution of the filter, and so
+there is no need to take care of the input -- it can be destroyed by the filter, and then deallocated.
+\verb$itk::FFTWRealToComplexConjugateImageFilter$ has been modified to use the input image directly without 
+intermediate buffer when \verb$ReleaseDataFlag$ is \verb$ON$. The detection of the \verb$ReleaseDataFlag$
+is a bit difficult during the pipeline 
+execution: this flag is always modified to \verb$OFF$ before the call to \verb$GenerateData()$ to avoid the 
+destruction by a minipipeline. In consequence, the detection of the \verb$ReleaseDataFlag$ is done at an 
+earlier stage in the pipeline execution, in \verb$UpdateOutputData()$, and stored in a member variable:
+\verb$m_CanUseDestructiveAlgorithm$.
+
+The later reuse of the plan, while sensible when not using the \verb$FFTW_ESTIMATE$ flag, is not much 
+interesting when it is used: the time needed to create the plan is very short -- way shorter than the
+FFT computation. Also, the number of threads and the image size are stored in the plan, and can't be 
+changed later, which doesn't fit well with ITK pipeline model. In the enhanced version, the plan is
+not kept between the filters execution, and so no intermediate output buffer are used, decreasing 
+significantly the needed memory.
+
+Another reason to create the internal buffer may be the need to align properly the image, to significantly
+improve the FFTW performance by using the SIMD operations. The buffer were simply allocated with the
+\verb$new$ operator, which is not enough to ensure a proper alignement
+on some plateforms, including Microsoft Windows and Linux.
+
 \subsubsection{Code safety}
+
+Only \verb$fftwf_execute(plan);$ and \verb$fftw_execute(plan);$ are thread safe in FFTW -- the other functions
+are not thread safe. In consequence, the call to all the other functions has been protected by a 
+global lock in the enhanced classes.
+
+Also, the check for the size of the image to know if the plan can be reused may not have been good 
+enough, because it was testing only the number of pixels and not the size on all the dimensions. 
+The plan is not kept between the execution, so this can not be a problem anymore.
 
@@
 
+At the end of the convolution, the image is cropped to return to the size of the input image. Again
+there is currently no option in ITK to do that fully in the pipeline model. In consequence, a new
+filter has been developped for this task.
+
+\verb$itk::RegionFromReferenceImageFilter$ is implemented as a subclass of \verb$itk::ExtractImageFilter$.
+This filter requires two inputs -- the first is the image to crop, the second is simply used to get 
+the region to extract in the first one. The reference image -- the second input -- is simply expected to
+be of type \verb$itk::ImageBase$ with the same dimension than the first one, and so it's type don't
+need to be specified in the template parameters.
+
 \subsection{FFTConvolutionImageFilter}
+
+\verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
+filters to make the FFT based convolution in a single and easy to use filter.
+It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
+user specified number of threads.
+
+It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTZeroPaddingImageFilter$ to let
+the user choose between the extra performance added by the padding and the extra memory cost.
 

### 2010-03-08 itkZeroFluxNeumannPadImageFilter.h
add ZeroFluxNeumannPadImageFilter
Ignore-this: 8d512e9c09686eb0a5e04b87ea0d1831

darcs-hash:20100308041443-2fc9d-fdcb69ce9fca9f57c4fd1c9fde6b2e49a7d27b66.gz

@@ -1,0 +1,130 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkZeroFluxNeumannPadImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009-04-25 12:27:20 $
+  Version:   $Revision: 1.19 $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkZeroFluxNeumannPadImageFilter_h
+#define __itkZeroFluxNeumannPadImageFilter_h
+
+#include "itkPadImageFilter.h"
+
+namespace itk
+{
+
+/** \class ZeroFluxNeumannPadImageFilter
+ * \brief Increase the image size by padding with a constant value.
+ *
+ * ZeroFluxNeumannPadImageFilter changes the output image region.  If the output
+ * image region is larger than the input image region, the extra pixels are
+ * filled in by a constant value.  The output image region must be specified.
+ *
+ * This filter is implemented as a multithreaded filter.  It provides a 
+ * ThreadedGenerateData() method for its implementation.
+ * 
+ * \ingroup GeometricTransforms
+ * \sa WrapPadImageFilter, MirrorPadImageFilter
+ */
+template <class TInputImage, class TOutputImage>
+class ITK_EXPORT ZeroFluxNeumannPadImageFilter:
+    public PadImageFilter<TInputImage,TOutputImage>
+{
+public:
+  /** Standard class typedefs. */
+  typedef ZeroFluxNeumannPadImageFilter             Self;
+  typedef PadImageFilter<TInputImage,TOutputImage>  Superclass;
+  typedef SmartPointer<Self>                        Pointer;
+  typedef SmartPointer<const Self>                  ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro(Self); 
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro(ZeroFluxNeumannPadImageFilter, PadImageFilter);
+
+  /** Typedef to describe the output image region type. */
[…]

### 2010-03-08 itkFFTZeroPaddingImageFilter.h
let FFTZeroPaddingImageFilter use different padding method - use Zero Flux Neumann by default
Ignore-this: 5f1e1cf8d8b8141c5ad607e92c801330

darcs-hash:20100308041545-2fc9d-5932d4bcffe3a8f8d4d7a8a55757df470cff853c.gz

@@ -121,6 +121,10 @@ public:
   itkGetConstMacro(GreatestPrimeFactor, int);
   itkSetMacro(GreatestPrimeFactor, int);
   
+  typedef enum { NO_PADDING=0, ZERO_FLUX_NEUMANN=1, ZERO=2, MIRROR=3, WRAP=4 } PadMethod;
+  itkGetConstMacro(PadMethod, int);
+  itkSetMacro(PadMethod, int);
+  
   /** Set the kernel image */
   void SetInputKernel(const InputKernelType *input)
     {
@@ -182,6 +186,7 @@ private:
   void operator=(const Self&); //purposely not implemented
 
   int m_GreatestPrimeFactor;
+  int m_PadMethod;
 
   bool isPrime( int n )
     {

### 2010-03-08 itkFFTPadImageFilter.h
rename FFTZeroPaddingImageFilter in FFTPadImageFilter
Ignore-this: bec34d16ecddd70f0237628375112690

darcs-hash:20100308042411-2fc9d-50250eed5620bbc951b34571b6a9e88345d05c81.gz

@@ -1,7 +1,7 @@
 /*=========================================================================
 
   Program:   Insight Segmentation & Registration Toolkit
-  Module:    $RCSfile: itkFFTZeroPaddingImageFilter.h,v $
+  Module:    $RCSfile: itkFFTPadImageFilter.h,v $
   Language:  C++
   Date:      $Date: 2009-01-28 18:14:36 $
   Version:   $Revision: 1.7 $
@@ -14,15 +14,15 @@
      PURPOSE.  See the above copyright notices for more information.
 
 =========================================================================*/
-#ifndef __itkFFTZeroPaddingImageFilter_h
-#define __itkFFTZeroPaddingImageFilter_h
+#ifndef __itkFFTPadImageFilter_h
+#define __itkFFTPadImageFilter_h
 
 #include "itkImageToImageFilter.h"
 #include "itkConceptChecking.h"
 
 namespace itk {
 
-/** \class FFTZeroPaddingImageFilter
+/** \class FFTPadImageFilter
  * \brief Pad two images with zeros to make them suitable for a convolution in the
  * frequency domain.
  *
@@ -45,12 +45,12 @@ namespace itk {
  */
 template<class TInputImage, class TInputKernel=TInputImage,
          class TOutputImage=TInputImage, class TOutputKernel=TInputKernel>
-class ITK_EXPORT FFTZeroPaddingImageFilter : 
+class ITK_EXPORT FFTPadImageFilter : 
     public ImageToImageFilter<TInputImage, TOutputImage>
 {
 public:
   /** Standard class typedefs. */
-  typedef FFTZeroPaddingImageFilter Self;
+  typedef FFTPadImageFilter Self;
 
   typedef ImageToImageFilter<TInputImage, TOutputImage> Superclass;
 
@@ -84,7 +84,7 @@ public:
   itkNewMacro(Self);  
 
   /** Runtime information support. */
-  itkTypeMacro(FFTZeroPaddingImageFilter, ImageToImageFilter);
+  itkTypeMacro(FFTPadImageFilter, ImageToImageFilter);
 
   /**
    * Set/Get whether the images must be padded to a size equal to a power of two.
@@ -169,8 +169,8 @@ public:
 
 
 protected:
-  FFTZeroPaddingImageFilter();
-  ~FFTZeroPaddingImageFilter() {};
+  FFTPadImageFilter();
+  ~FFTPadImageFilter() {};
[…]

### 2010-03-08 itkFFTConvolutionImageFilter.h
expose PadMethod in FFTConvolutionImageFilter
Ignore-this: 8f0d0d85eaf62cae066975782cbe3936

darcs-hash:20100308053321-2fc9d-1e47a4342dd65b42b8f19c11de304d889fbf4fb4.gz

@@ -130,6 +130,13 @@ public:
   itkGetConstMacro(GreatestPrimeFactor, int);
   itkSetMacro(GreatestPrimeFactor, int);
   
+  /**
+   * Set/Get the padding method.
+   */
+  typedef enum { NO_PADDING=0, ZERO_FLUX_NEUMANN=1, ZERO=2, MIRROR=3, WRAP=4 } PadMethod;
+  itkGetConstMacro(PadMethod, int);
+  itkSetMacro(PadMethod, int);
+  
 #ifdef ITK_USE_CONCEPT_CHECKING
   /** Begin concept checking */
   itkConceptMacro(InputHasPixelTraitsCheck,
@@ -157,6 +164,7 @@ private:
   void operator=(const Self&); //purposely not implemented
 
   int m_GreatestPrimeFactor;
+  int m_PadMethod;
 
 }; // end of class
 

### 2010-03-08 itkFFTPadImageFilter.h

@@ -121,6 +121,9 @@ public:
   itkGetConstMacro(GreatestPrimeFactor, int);
   itkSetMacro(GreatestPrimeFactor, int);
   
+  /**
+   * Set/Get the padding method.
+   */
   typedef enum { NO_PADDING=0, ZERO_FLUX_NEUMANN=1, ZERO=2, MIRROR=3, WRAP=4 } PadMethod;
   itkGetConstMacro(PadMethod, int);
   itkSetMacro(PadMethod, int);

### 2010-03-08 itkFFTConvolutionImageFilter.h
add a Normalize option
Ignore-this: 7dc6bf95ff4767610ac5627fe51cb0c4

darcs-hash:20100308222047-2fc9d-7adca213455520b0a9534097e1c46a21dcc79e70.gz

@@ -137,6 +137,14 @@ public:
   itkGetConstMacro(PadMethod, int);
   itkSetMacro(PadMethod, int);
   
+  /**
+   * Set/Get whether the kernel should be normalized to one or not.
+   * Default is true.
+   */
+  itkGetConstMacro(Normalize, bool);
+  itkSetMacro(Normalize, bool);
+  itkBooleanMacro(Normalize);
+  
 #ifdef ITK_USE_CONCEPT_CHECKING
   /** Begin concept checking */
   itkConceptMacro(InputHasPixelTraitsCheck,
@@ -163,8 +171,9 @@ private:
   FFTConvolutionImageFilter(const Self&); //purposely not implemented
   void operator=(const Self&); //purposely not implemented
 
-  int m_GreatestPrimeFactor;
-  int m_PadMethod;
+  int  m_GreatestPrimeFactor;
+  int  m_PadMethod;
+  bool m_Normalize;
 
 }; // end of class
 

### 2010-03-08 itkZeroFluxNeumannPadImageFilter.h
better doc
Ignore-this: 6da49af9a9bb646c250517f4f2e2921a

darcs-hash:20100308222544-2fc9d-3adbdff800553b0d96dd8ed80a823fccba15e569.gz

@@ -25,15 +25,30 @@ namespace itk
 /** \class ZeroFluxNeumannPadImageFilter
  * \brief Increase the image size by padding with a constant value.
  *
- * ZeroFluxNeumannPadImageFilter changes the output image region.  If the output
- * image region is larger than the input image region, the extra pixels are
- * filled in by a constant value.  The output image region must be specified.
+ * A filter which extends the image size and fill the missing pixels
+ * according to a Neumann boundary condition where first,
+ * upwind derivatives on the boundary are zero.  This is a useful condition 
+ * in solving some classes of differential equations. 
  *
- * This filter is implemented as a multithreaded filter.  It provides a 
- * ThreadedGenerateData() method for its implementation.
+ * For example, invoking this filter on an image with a corner like:
+ * \code
+ *               * * * * * * * 
+ *               * * * * * * *
+ *               * * 1 2 3 4 5  (where * denotes pixels that lie 
+ *               * * 3 3 5 5 6          outside of the image boundary)
+ *               * * 4 4 6 7 8
+ * \endcode
+ * returns the following padded image:
+ * \code
+ *               1 1 1 2 3 4 5
+ *               1 1 1 2 3 4 5
+ *               1 1 1 2 3 4 5
+ *               3 3 3 3 5 5 6   (note the corner values)
+ *               4 4 4 4 6 7 8
+ * \endcode
  * 
  * \ingroup GeometricTransforms
- * \sa WrapPadImageFilter, MirrorPadImageFilter
+ * \sa WrapPadImageFilter, MirrorPadImageFilter, ConstantPadImageFilter, ZeroFluxNeumannBoundaryCondition
  */
 template <class TInputImage, class TOutputImage>
 class ITK_EXPORT ZeroFluxNeumannPadImageFilter:

### 2010-03-09 article/Article.tex
article continued
Ignore-this: 10cd3a5e71450026f96dd7c30dc2cf63

darcs-hash:20100309014443-2fc9d-edfd21b5f498036bd9e9a7a142bb1f7471c20687.gz

@@
 
-\subsection{Zero padding}
+\subsection{Padding}
 
-The images will be padded with zeros for several reasons:
+The image to convolve and the kernel will be padded with for several reasons:
 \begin{itemize}
@@
 
+The kernel is always padded with zeros, but the image to convolve can be padded in several ways:
+\begin{itemize}
+ \item zeros -- while simple, this method may show darkened border in the convolved image;
+ \item zero flux Neumann -- this method extends the pixels on the image border in the padded zone, and
+       is quite efficient in keeping the border effect low;
+ \item mirror -- the image is padded with a mirror of the image. This is also a quite good way to
+       keep the border effect low;
+ \item wrap -- the opposite border of the image is copied in the padded zone. This may be useful to
+       keep the circular behavior of the FFT, while still padding to fit the size requirement of the
+       FFT implementation.
+\end{itemize}
+The result of those padding can be seen in figure \ref{fig:padding}.
+
+\begin{figure}[htbp]
+\begin{center}
+\subfigure[Zero]{\includegraphics[scale=0.4]{pad-image-2}}
+\subfigure[Zero flux Neumann]{\includegraphics[scale=0.4]{pad-image-zero-flux}}
+\subfigure[Mirror]{\includegraphics[scale=0.4]{pad-image-mirror}}
+\subfigure[Wrap]{\includegraphics[scale=0.4]{pad-image-wrap}}
+\subfigure[Input image]{\includegraphics[scale=0.4]{cthead1}}
+\caption{Padding a 256 $\times$ 256 image to a 512 $\times$ 512 one in several ways.\label{fig:padding}}
+\end{center}
+\end{figure}
+
 \subsection{Normalization}
@@
 convolved image, but flipped on all the axes. To get the same behavior when computing the product
-in the frequency domain, the kernel image must be flipped prior to the FFT.
+in the frequency domain, the kernel image must be flipped prior to the FFT. See figure \ref{fig:kernel-modifications}.
+
+\begin{figure}[htbp]
+\begin{center}
+\subfigure[Input image (scaled 20 times)]{\includegraphics[scale=8]{kernel}}
+\subfigure[Zero padded]{\includegraphics[scale=0.4]{pad-kernel-2}}
+\subfigure[Zero padded and flipped]{\includegraphics[scale=0.4]{kernel-pad-flip}}
+\subfigure[Zero padded, flipped and shifted]{\includegraphics[scale=0.4]{kernel-pad-flip-shift}}
+\caption{Kernel modifications in order to perform the FFT based convolution. Normalization is
+only a change in intensity, and thus is not shown here.\label{fig:kernel-modifications}}
+\end{center}
+\end{figure}
 
@@
 usual for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
-of the image as expeced by the FFT.
+of the image as expeced by the FFT. See figure \ref{fig:kernel-modifications}.
 
@@
 The convolved image in the frequency domain is transformed to the space domain with an inverse Fourier
-transform.
+transform. See figure \ref{fig:image-modifications}.
+
+\begin{figure}[htbp]
+\begin{center}
+\subfigure[Input image]{\includegraphics[scale=0.4]{cthead1}}
+\subfigure[Zero flux Neumann padded]{\includegraphics[scale=0.4]{pad-image-zero-flux}}
+\subfigure[Convolved, still padded]{\includegraphics[scale=0.4]{image-pad-convolved}}
+\subfigure[Cropped to the input image size]{\includegraphics[scale=0.4]{fftconv}}
+\caption{The modifications applied to the input image.\label{fig:image-modifications}}
+\end{center}
+\end{figure}
 
@@
 
-The image is cropped to fit the size of the input image.
+The image is cropped to fit the size of the input image. See figure \ref{fig:image-modifications}.
 
@@
 
-\subsection{FFTZeroPaddingImageFilter}
+\subsection{ZeroFluxNeumannPadImageFilter}
+
+The zero flux Neumann padding filter was not implemented in ITK. It has been implemented by modifying
+the \verb$ConstantPadImageFilter$ code.
+
+\subsection{FFTPadImageFilter}
 
@@
 image size used to pad both the input image and the kernel.
-All the logic is implemented in itk::FFTZeroPaddingImageFilter.
+All the logic is implemented in \verb$itk::FFTPadImageFilter$.
 
@@
 
-The image to convolve is simply padded with zeros. The kernel, however, requires a little more work to be
+The image to convolve is simply padded with one of the available method: zeros, zero flux Neumann, mirror
+or wrap.
+
+The kernel, however, requires a little more work to be
 centered in the padded image, and to make its region match the one of the padded image to convolve.
-The padding and the region changes are implemented as a minipipeline of itk::ConstantPadImageFilter and
-itk::ChangeInformationImageFilter.
+The padding and the region changes are implemented as a minipipeline of \verb$itk::ConstantPadImageFilter$ and
+\verb$itk::ChangeInformationImageFilter$.
 
-itk::FFTZeroPaddingImageFilter is also able to extend the padded region in order to enhance the FFT
+\verb$itk::FFTPadImageFilter$ is also able to extend the padded region in order to enhance the FFT
 performance, or to make it possible when the FFT possible when it require a size which is a power of
@@
 \end{pseudocode}
+
+\verb$FFTPadImageFilter$ can also be used with a single input. This is useful to be able to run a FFT
+on an image with a size which is not a power of 2 with the vnl implementation, or to enhance the performance
+of the FFTW implementation, without necessairly doing a convolution.
 
@@
 
+\subsubsection{Still to be done}
+
+There are still some missing features in the FFTW filters. The ones which comes to mind are:
+\begin{itemize}
+ \item let the user create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
+\verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a
+later run;
+ \item provide an API to save the FFTW wisdoms.
+\end{itemize}
+
 \subsection{RegionFromReferenceImageFilter}
@@
 \verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
-filters to make the FFT based convolution in a single and easy to use filter.
+filters to make the FFT based convolution in a single and easy to use filter. See figure \ref{fig:conv-pipeline}
+for an overview of the minipipeline.
+\begin{figure}[htbp]
+\centering
+\includegraphics[scale=0.5]{conv}
+\caption{Internal pipeline of FFTConvolutionImageFilter.\label{fig:conv-pipeline}}
+\end{figure}
+
 It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
@@
 
-It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTZeroPaddingImageFilter$ to let
-the user choose between the extra performance added by the padding and the extra memory cost.
+It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTPadImageFilter$ to let
+the user choose between the extra performance added by the padding and the extra memory cost, and
+\verb$PadMethod$ from the same filter to let the user choose the padding method -- zero flux Neumann
+is used by default.
+
+Finally, the \verb$Normalize$ attribute let the user choose if the kernel must be normalized to one
+or not.
 
@@
 \label{perf:gpf}
+As visible in figure \ref{fig:perf-fftw3}, the FFTW library performs quite differently depending on the
+size of the input. A good way to get the best performance seems to be to use the sizes with the greatest
+prime factor smaller or equal to 13.
+
+Most of the time, and despite the usual advice to use a size which is a power of two,
+those sizes perform significantly better than the closest sizes which are a power of two, probably
+because of the smaller impact of the number of pixels in the image.
+
+\begin{figure}[htbp]
+\begin{center}
+\subfigure[Execution time for an image of size $dim^{3}$.]{\includegraphics[scale=0.7]{fftw3-time-dim}}
+\subfigure[Effect of the image greatest prime factor of the image size (gpf) on the number of pixels
+ processed per seconds (ops).]{\includegraphics[scale=0.7]{fftw3-ops-gpf}}
+\caption{Effect of the image size on the FFTW performance. Red dots are the sizes with a greatest prime
+factor smaller or equal to 13. The grey dots are the ones with a greatest prime factor greater than 13.
+\label{fig:perf-fftw3}}
+\end{center}
+\end{figure}
 
@@
 
-% \url{http://www.itk.org}
-% \code{Insight/Documentation/Style.pdf}
-% \section{Principles of Solar Spot Detection}
-% \cite{ITKSoftwareGuide}.
-% \doxygen{ImageToImageFilter}
+\section{Conclusion}
 
-% \small \begin{verbatim}
-% \end{verbatim} \normalsize
+This contribution provides with an efficient implementation of the convolution transform for large kernels.
+The FFTW filters implementations has been significantly enhanced, and a set of reusable tools is provided.
 
-% The {itemize} environment uses a bullet for each \item.  If you want the 
-% \item's numbered, use the {enumerate} environment instead.
-% \begin{itemize}
-%   \item  Insight Toolkit 2.4.
-%   \item  CMake 2.2
-% \end{itemize}
-
-% \ref{cthead1}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{cthead1}
-\caption{The input image.\label{cthead1}}
-\end{figure}
+Some other work may be done on this basis -- an non exhaustive list may be:
+\begin{itemize}
+ \item a binary erosion/dilation based on FFT;
+ \item a set of deconvolution filters;
+ \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm.
+\end{itemize}
 

### 2010-03-10 article/Article.tex
update article and style with the new IJ style
Ignore-this: 83f17015111d18b0752b208ac8ce75e9

darcs-hash:20100309042535-2fc9d-ead501d6d8029574558b94b7e7af0de3dd9982e6.gz
article continued
Ignore-this: 6b8dfb502f3001b366ff081bda9cfb0c

darcs-hash:20100310000931-2fc9d-9873e486bf49c07fcf8f97391d31823a6eb71e05.gz

@@
 
+% 
+% NOTE: This is the last number of the "handle" URL that 
+% The Insight Journal assigns to your paper as part of the
+% submission process. Please replace the number "1338" with
+% the actual handle number that you get assigned.
+%
+\newcommand{\IJhandlerIDnumber}{}
+
 % Increment the release number whenever significant changes are made.
@@
 \begin{document}
+
+%
+% Add hyperlink to the web location and license of the paper.
+% The argument of this command is the handler identifier given
+% by the Insight Journal to this paper.
+% 
+\IJhandlefooter{\IJhandlerIDnumber}
+
 \maketitle
@@
 \end{abstract}
+
+\IJhandlenote{\IJhandlerIDnumber}
 
@@
 
-\subsubsection{Still to be done}
-
-There are still some missing features in the FFTW filters. The ones which comes to mind are:
-\begin{itemize}
- \item let the user create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
-\verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a
-later run;
- \item provide an API to save the FFTW wisdoms.
-\end{itemize}
-
 \subsection{RegionFromReferenceImageFilter}
@@
 
+All the performance tests are run on a workstation with two Intel(R) Xeon(R) CPU X5570 at 2.93GHz
+and 24 GB of RAM and running Mandriva Linux 2009.1.
+
 \subsection{Kernel size}
+
+As visible in figure \ref{fig:perf-kernel}, the kernel size as a little effect on the execution time
+for the FFT based convolution, but a quite big impact on the normal one.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{perf2D}
+\caption{The effect of the kernel size on the execution time. The input image size is $256 \times 256$.
+\label{fig:perf-kernel}}
+\end{figure}
+
+The drop in performance at some times seems to be due to a not so good plan generated by FFTW with \verb$FFTW_ESTIMATE$.
+Using FFTW's wisdom feature can almost remove that problem. An integration of wisdom in ITK's FFTW
+filters will be proposed in a later contribution.
 
@@
 those sizes perform significantly better than the closest sizes which are a power of two, probably
-because of the smaller impact of the number of pixels in the image.
+because the performance gain per pixel doesn't compensate the higher number of pixels in the image.
 
@@
 
+As visible in figure \ref{fig:perf-kernel}, the FFT based convolution and the normal one are scaling
+quite well with the number of threads.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{perf_threads}
+\caption{The effect of the number of threads on the execution time. The input image size is
+$371 \times 371 \times 34$, and the kernel size is two.
+\label{fig:perf-kernel}}
+\end{figure}
+
+\section{Wrapping}
+
+All the new filters have been wrapped using WrapITK.
+
+\section{Development version}
+
+A development version is available in a darcs repository at
+\url{http://mima2.jouy.inra.fr/darcs/contrib-itk/fftconv/}.
+
 \section{Conclusion}
@@
  \item a set of deconvolution filters;
- \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm.
+ \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm;
+ \item let the user create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
+       \verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a
+       later run;
+ \item integrate FFTW's wisdom to improve FFT performance in ITK.
 \end{itemize}

### 2010-03-10 article/Article.tex
article continued
Ignore-this: 3cb26f566f0ce5f1a6ab5a859d1c8dae

darcs-hash:20100310002103-2fc9d-2de7876faf00dd5e223363042031a1eaf6e4c808.gz

@@
 %
-\newcommand{\IJhandlerIDnumber}{}
+\newcommand{\IJhandlerIDnumber}{3154}
 
@@
 This paper provides such a filter, as well as a detailed description of the implementation
-choices and a comparison with the "simple" itk::ConvolutionImageFilter.
+choices and a performance comparison with the "simple" itk::ConvolutionImageFilter.
 

### 2010-03-19 article/Article.tex
remove inra and add Acknowledgments
Ignore-this: 6f02b0970b6c7761505bbb9d0c0dbf03

darcs-hash:20100319034836-2fc9d-77547132d563ec7f2a136ec4483b52a966bedf38.gz

@@
 % snail-mail address if you like.
-\author{Ga\"etan Lehmann{$^1$}}
-\authoraddress{{$^1$}INRA, UMR 1198; ENVA; CNRS, FRE 2857, Biologie du
-D\'eveloppement et 
-Reproduction, Jouy en Josas, F-78350, France.}
+\author{Ga\"etan Lehmann}
+% \authoraddress{}
 
@@
 
+\section{Acknowledgments}
+
+Unlike most of my previous contributions, this work has {\em not} been supported by
+the french National Institute for Agricultural Research (INRA) -- my employer.
+I'd like to thanks the INRA anyway, and more specifically the MIMA2 facility, to
+have provided the computer ressources used for this work.
+
 \section{Conclusion}

### 2010-03-25 article/Article.tex
Dan White's corrections
Ignore-this: f7ff0d7ab5c511de8c08eab081b91758

darcs-hash:20100325235045-2fc9d-ff085e090a08e44d287eadffa1b8a1e2e816f3e0.gz

@@
 The Fourier transform of the convolution of two images is equal to the product of their
-Fourier transform. With this definition, it is possible to create a convolution filter
+Fourier transforms. With this definition, it is possible to create a convolution filter
 based on the Fast Fourier Transform (FFT). The interesting complexity characteristics of
-this transform gives a very efficient convolution filter for large kernel images.
+this transform give a very efficient convolution filter for large kernel images.
 
@@
 
-The convolution of two image is a very cpu demanding task, with a complexity of $O(N \times M)$ where $N$ is
-the number of pixel of the first image and $M$ the number of pixels of the second. Usually, the size
-of one of the image is a lot smaller than the one of the other -- the smaller is often called the
-kernel. The quite small size of the kernel makes the convolution computable in a very usable time,
-but when the size of the kernel grows, the computation time quickly becomes impraticable.
+The convolution of two images is a very cpu demanding task, with a complexity of $O(N \times M)$ where $N$ is
+the number of pixels of the first image and $M$ the number of pixels of the second. Usually, the size
+of one of the images is much smaller than the one of the other; the smaller image is often called the
+kernel. The relatively small size of the kernel makes the convolution computable in a reasonable time,
+but when the size of the kernel grows, the computation times quickly become impractical.
 
@@
 The cost of the Fourier Transform
-can be quite high however -- that's why the FFT based convolution is more efficient only for big
+can be quite high however -- that's why the FFT based convolution is more efficient only for large
 kernel images.
@@
 
-The FFT based convolution requires several step to be performed. They are detailed one by one.
+The FFT based convolution requires several steps to be performed. They are detailed here one by one.
 
@@
 
-The image to convolve and the kernel will be padded with for several reasons:
+The image to convolve and the kernel image will be padded for several reasons:
 \begin{itemize}
-  \item to make their size match. This is required to perform the multiplication in the frequency
+  \item to make their sizes match. This is required to perform multiplication in the frequency
   domain.
-  \item  to avoid the border effects. Because the FFT considere the image as a cyclic signal, the
-  image must be padded to avoid the border effects. The final image size for all the dimensions
-  must be at least $P+Q-1$, where $P$ is the size of the first image on a dimension and $Q$ the size
-  of the other image on the same dimension.
-  \item to make the FFT possible, or to enhance its performance. Some FFT implementations, like the
-  VNL one, can only be run on an image where the size on all the dimensions are a power of two.
-  Some other implementations, like FFTW's one, are performing very differently depending on the
+  \item  to avoid border effects. Because the FFT considers the image as a cyclic signal, the
+  image must be padded to avoid border effects. The final image size for all the dimensions
+  must be at least $P+Q-1$, where $P$ is the size of the first image in a dimension and $Q$ the size
+  of the other image in the same dimension.
+  \item to make the FFT possible, or to enhance its performance. Some FFT implementations, like that in
+  VNL, can only be run on an image where the size of all the dimensions are powers of two.
+  Some other implementations, like that of FFTW, perform very differently depending on the
   size of the image. Adding a few more pixels on the border of the image can lead them to perform
-  a lot better, as shown in chapter \ref{perf:gpf}.
+  significantly better, as shown in chapter \ref{perf:gpf}.
 \end{itemize}
@@
 \begin{itemize}
- \item zeros -- while simple, this method may show darkened border in the convolved image;
+ \item zeros -- while simple, this method may produce a darkened border in the convolved image;
  \item zero flux Neumann -- this method extends the pixels on the image border in the padded zone, and
        is quite efficient in keeping the border effect low;
- \item mirror -- the image is padded with a mirror of the image. This is also a quite good way to
+ \item mirror -- the image is padded with a mirror of the image. This is also a good way to
        keep the border effect low;
@@
 \end{itemize}
-The result of those padding can be seen in figure \ref{fig:padding}.
+The result of these paddings can be seen in figure \ref{fig:padding}.
 
@@
 
-To preserve the intensity of the pixel in the convolved image, the kernel must be normalized to one
--- the sum of all its pixel is one.
+To preserve the intensity of the pixels in the convolved image, the kernel must be normalized to one
+-- the sum of all its pixels is one.
 
@@
 
-The convolution is a flipping transform by nature: the shape of the kernel image can be found in the
+The convolution is by nature a flipping transform: the shape of the kernel image can be found in the
 convolved image, but flipped on all the axes. To get the same behavior when computing the product
@@
 
-The convolution is usually done with a centered kernel. If the kernel is not centered, the input image
+The convolution is usually performed with a centered kernel. If the kernel is not centered, the input image
 is shifted in the convolved image, by the same shift than the shift of the kernel. However, for the
-FFT, the center pixel is the one in the corner, not the one in the center of the image, as is it 
-usual for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
+FFT, the center pixel is the one in the corner, not the one in the center of the image, as is intuitive
+for a human. To avoid a shift in the convolved image, the pixels must be shifted to the corner
 of the image as expeced by the FFT. See figure \ref{fig:kernel-modifications}.
@@
 The convolution is performed in the frequency domain, simply by computing the product of the FFT of
-the image and of the FFT on the kernel pixel wise.
+the image and the FFT of the kernel, pixel wise.
 
@@
 
-The normalization to one cannot be done without going outside of the ITK pipeline model at this time.
+The normalization to unity cannot be done without going outside of the ITK pipeline model at this time.
 itk::NormalizeToConstantImageFilter has been added for this task.
@@
 
-This filter is in charge of padding both images. It requires two input -- the image to convolve and
+This filter is in charge of padding both images. It requires two inputs: the image to convolve and
 the kernel. The pad size is computed according to both image sizes.
@@
 The kernel, however, requires a little more work to be
-centered in the padded image, and to make its region match the one of the padded image to convolve.
+centered in the padded image, and to make its region match that of the padded image to be convolved.
 The padding and the region changes are implemented as a minipipeline of \verb$itk::ConstantPadImageFilter$ and
@@
 \verb$itk::FFTPadImageFilter$ is also able to extend the padded region in order to enhance the FFT
-performance, or to make it possible when the FFT possible when it require a size which is a power of
-two. FFTW is able to work on any image size, but due to the algorithms useds, produce significantly
-more performant result with some specific sizes. The two main reasons are:
+performance, or to make it possible when the FFT method requires a size which is a power of
+two. FFTW is able to work on any image size, but due to the algorithms used, it is significantly faster
+with some specific image sizes. The two main reasons are:
 \begin{itemize}
@@
   Cooley-Tukey algorithm;
-  \item the hard coded loop unrolling for the size which are prime numbers up to thirteen.
+  \item the hard coded loop unrolling for the sizes which are prime numbers up to thirteen.
 \end{itemize}
-For these reasons, FFTW performs better when the size on each dimension has its greatest prime
+For these reasons, FFTW performs better when the size of each dimension has its greatest prime
 factor smaller or equal to thirteen.
-The greatest prime factor can be simply computed with the algorithm described in algorithm \ref{greatestPrimeFactor}.
+The greatest prime factor can be computed easily with the algorithm described in algorithm \ref{greatestPrimeFactor}.
 
@@
 
-\verb$FFTPadImageFilter$ can also be used with a single input. This is useful to be able to run a FFT
+\verb$FFTPadImageFilter$ can also be used with a single input. This is useful for running a FFT
 on an image with a size which is not a power of 2 with the vnl implementation, or to enhance the performance
-of the FFTW implementation, without necessairly doing a convolution.
+of the FFTW implementation, without necessairly doing a convolution later.
 
@@
 \verb$itk::fftw::Proxy$ has been modified to provide this feature: an extra \verb$thread$ parameter,
-which defaults
-to $1$ to keep the backward compatibility, has been added to the method in charge of creating the plans.
-Then the \verb$itk::FFTWComplexConjugateToRealImageFilter$ and
+has been added to the method in charge of creating the plans. This defaults
+to $1$ to preserve backwards compatibility.
+Further, the \verb$itk::FFTWComplexConjugateToRealImageFilter$ and
 \verb$itk::FFTWRealToComplexConjugateImageFilter$ have
@@
 FFTW filters were using a lot of memory, because they were allocating an internal buffer for the input 
-image, and an other one for the output image. Worth, these buffers were not deallocated at the end of
+image, and another for the output image. Worse, these buffers were not deallocated at the end of
 the execution filter, keeping the memory usage very high even when the execution has completed.
 
-I guess the main reasons to do that were:
+One supposes the main reasons to do that were:
 \begin{itemize}
   \item avoid the destruction of the input. The creation of the FFTW plan, and the execution of FFTW 
-  can destroy the input. This is most of the time not acceptable in ITK.
-  \item reuse the FFTW plan for a latter run. The plan include the memory location of the input and output
-  buffers, so the buffers can't be deallocated, as it may be the case with the filter's input and output images.
+  can destroy the input. This is not acceptable most of the time in ITK.
+  \item reuse the FFTW plan for a latter run. The plan includes the memory location of the input and output
+  buffers, so the buffers can't be deallocated, as may be the case with the filter's input and output images.
 \end{itemize}
@@
 the input buffer.
-The case is a bit
+The case is
 more difficult for the complex to real transform: FFTW does not provide any algorithm able to preserve
-the input, so the input must be copied to an internal buffer. There is a case however, when we don't
+the input, so the input must be copied to an internal buffer. There is a case however, when one does not
 care about destroying the input in ITK: when the \verb$ReleaseDataFlag$ is \verb$ON$ on the input of the filter.
 In that case, the input will be deallocated right after the end of the execution of the filter, and so
-there is no need to take care of the input -- it can be destroyed by the filter, and then deallocated.
-\verb$itk::FFTWRealToComplexConjugateImageFilter$ has been modified to use the input image directly without 
+there is no need to take care of the input: it can be destroyed by the filter, and then deallocated.
+\verb$itk::FFTWRealToComplexConjugateImageFilter$ has been modified to use the input image directly without an
 intermediate buffer when \verb$ReleaseDataFlag$ is \verb$ON$. The detection of the \verb$ReleaseDataFlag$
-is a bit difficult during the pipeline 
-execution: this flag is always modified to \verb$OFF$ before the call to \verb$GenerateData()$ to avoid the 
+is somewhat difficult during the pipeline 
+execution: this flag is always modified to \verb$OFF$ before the call to \verb$GenerateData()$ to avoid
 destruction by a minipipeline. In consequence, the detection of the \verb$ReleaseDataFlag$ is done at an 
@@
 
-The later reuse of the plan, while sensible when not using the \verb$FFTW_ESTIMATE$ flag, is not much 
-interesting when it is used: the time needed to create the plan is very short -- way shorter than the
+The later reuse of the plan, while sensible when not using the \verb$FFTW_ESTIMATE$ flag, is of little practical use
+since the time needed to create the plan is very short; much shorter than the
 FFT computation. Also, the number of threads and the image size are stored in the plan, and can't be 
 changed later, which doesn't fit well with ITK pipeline model. In the enhanced version, the plan is
-not kept between the filters execution, and so no intermediate output buffer are used, decreasing 
-significantly the needed memory.
+not kept between the filters execution, and so no intermediate output buffers are used,
+significantly decreasing memory requirements.
 
-Another reason to create the internal buffer may be the need to align properly the image, to significantly
-improve the FFTW performance by using the SIMD operations. The buffer were simply allocated with the
-\verb$new$ operator, which is not enough to ensure a proper alignement
+Another reason to create the internal buffer may be the need to properly align the image, to significantly
+improve the FFTW performance by using SIMD operations of the CPU. The buffers were simply allocated with the
+\verb$new$ operator, which is not enough to ensure a proper alignment
 on some plateforms, including Microsoft Windows and Linux.
@@
 Only \verb$fftwf_execute(plan);$ and \verb$fftw_execute(plan);$ are thread safe in FFTW -- the other functions
-are not thread safe. In consequence, the call to all the other functions has been protected by a 
+are not thread safe. As a consequence, calls to all the other functions have been protected by a 
 global lock in the enhanced classes.
 
-Also, the check for the size of the image to know if the plan can be reused may not have been good 
-enough, because it was testing only the number of pixels and not the size on all the dimensions. 
-The plan is not kept between the execution, so this can not be a problem anymore.
+Also, image size checking to know if the plan can be reused may not have been safe,
+since it was testing the number of pixels rather than the size of all the dimensions. 
+With the changes already made in this classes for other reasons, the plan is no longer kept
+between successive executions, so this is no longer a problem.
 
@@
 
-At the end of the convolution, the image is cropped to return to the size of the input image. Again
-there is currently no option in ITK to do that fully in the pipeline model. In consequence, a new
+At the end of the convolution, the image is cropped to return it to the size of the input image. Again,
+there is currently no option in ITK to do that fully within the pipeline model. Consequently, a new
 filter has been developped for this task.
@@
 This filter requires two inputs -- the first is the image to crop, the second is simply used to get 
-the region to extract in the first one. The reference image -- the second input -- is simply expected to
-be of type \verb$itk::ImageBase$ with the same dimension than the first one, and so it's type don't
+the region to extract in the first one. The reference image, the second input, is simply expected to
+be of type \verb$itk::ImageBase$ with the same dimensionality as the first, and so it's type does not
 need to be specified in the template parameters.
@@
 \verb$itk::FFTConvolutionImageFilter$ is a convenient minipipeline filter which groups all the required
-filters to make the FFT based convolution in a single and easy to use filter. See figure \ref{fig:conv-pipeline}
+filters to perform the FFT based convolution in a single, easy to use filter. See figure \ref{fig:conv-pipeline}
 for an overview of the minipipeline.
@@
 
-It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagate the
+It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagates the
 user specified number of threads.
 
-It also expose the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTPadImageFilter$ to let
+It also exposes the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTPadImageFilter$ to let
 the user choose between the extra performance added by the padding and the extra memory cost, and
-\verb$PadMethod$ from the same filter to let the user choose the padding method -- zero flux Neumann
+\verb$PadMethod$ from the same filter to let the user choose the padding method: Zero flux Neumann
 is used by default.
 
-Finally, the \verb$Normalize$ attribute let the user choose if the kernel must be normalized to one
+Finally, the \verb$Normalize$ attribute lets the user choose if the kernel must be normalized to one
 or not.
@@
 
-All the performance tests are run on a workstation with two Intel(R) Xeon(R) CPU X5570 at 2.93GHz
-and 24 GB of RAM and running Mandriva Linux 2009.1.
+All the performance tests were run on a workstation with two Intel(R) Xeon(R) CPU X5570 at 2.93GHz
+and 24 GB of RAM and running Mandriva Linux 2009.1. This system has 8 cores and is able to physically
+run 16 threads at the same time.
 
@@
 As visible in figure \ref{fig:perf-kernel}, the kernel size as a little effect on the execution time
-for the FFT based convolution, but a quite big impact on the normal one.
+for the FFT based convolution, but a quite big impact on the spatial domain convolution method.
 
@@
 
-The drop in performance at some times seems to be due to a not so good plan generated by FFTW with \verb$FFTW_ESTIMATE$.
+The drop in performance at some times seems to be due to a suboptimal plan generated by FFTW with \verb$FFTW_ESTIMATE$.
 Using FFTW's wisdom feature can almost remove that problem. An integration of wisdom in ITK's FFTW
@@
 As visible in figure \ref{fig:perf-fftw3}, the FFTW library performs quite differently depending on the
-size of the input. A good way to get the best performance seems to be to use the sizes with the greatest
+size of the input. A good way to optimise performance appears to be use of image sizes having a greatest
 prime factor smaller or equal to 13.
@@
 those sizes perform significantly better than the closest sizes which are a power of two, probably
-because the performance gain per pixel doesn't compensate the higher number of pixels in the image.
+because the performance gain per pixel doesn't compensate for the higher number of pixels in the image.
 
@@
 
-As visible in figure \ref{fig:perf-kernel}, the FFT based convolution and the normal one are scaling
-quite well with the number of threads.
+As is visible in figure \ref{fig:perf-kernel}, the FFT based and spatial domain convolutions both scale
+well with the number of threads.
 
@@
 the french National Institute for Agricultural Research (INRA) -- my employer.
-I'd like to thanks the INRA anyway, and more specifically the MIMA2 facility, to
-have provided the computer ressources used for this work.
+I'd like to thank the INRA anyway, and more specifically the MIMA2 facility, who
+provided the computer ressources used for this work.
+
+I'm grateful to Dan White for its corrections and useful comments on this article.
 
@@
 
-This contribution provides with an efficient implementation of the convolution transform for large kernels.
-The FFTW filters implementations has been significantly enhanced, and a set of reusable tools is provided.
+This contribution provides an efficient implementation of the convolution transform for large kernels.
+The FFTW filter implementations have been significantly enhanced, and a set of reusable tools is provided.
 
-Some other work may be done on this basis -- an non exhaustive list may be:
+Some other work may be done on this basis. An non exhaustive list includes:
 \begin{itemize}
- \item a binary erosion/dilation based on FFT;
+ \item binary erosion/dilation filters based on FFT;
  \item a set of deconvolution filters;
  \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm;
- \item let the user create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
+ \item allow the user to create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
        \verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a

### 2010-07-21 article/Article.tex
updated article
Ignore-this: edd347be772d143645f2e83076229454

darcs-hash:20100721124622-2fc9d-44b31b30d2baf85f62059a53e0069abad263685e.gz

@@
 \author{Ga\"etan Lehmann}
-% \authoraddress{}
+\authoraddress{{$^1$}INRA, UMR 1198; ENVA; CNRS, FRE 2857, Biologie du
+D\'eveloppement et 
+Reproduction, Jouy en Josas, F-78350, France.}
 
@@
 \end{figure}
+
+\subsection{Clipping}
+
+Because of the numerical precision used in the FFT transforms, the values can sometime be greater or
+smaller than the possible values of the output image.
+The values are clipped to fit in the range of possible values of the output image.
 
@@
 
+\subsubsection{Plan optimization}
+
+The ITK classes have been modified in order to create optimized FFTW plan. Because the computation
+of the optimized plan can take a lot of time, the default is still \verb$ESTIMATE$.
+
+The optimization level can be changed globally only at this time, with the method
+\verb$itk::FFTWLock::SetGlobalOptimizationLevel()$, or by setting the environment variable
+\verb$ITK_FFTW_OPTIMIZATION_LEVEL$.
+
+\subsubsection{Wisdom integration}
+
+The optimized plans are automatically loaded and saved from/in the files \verb$~/itkwisdomf$ and \verb$~/itkwisdom$,
+or in the file name stored in the environment variable \verb$ITK_FFTW_WISDOM _FILE_BASE_NAME$ if this
+variable exists.
+They are saved automatically when the program exits, and can also be saved with some dedicated methods
+in \verb$itk::FFTWLock$.
+
+\subsection{ClipImageFilter}
+
+\verb$itk::ClipImageFilter$ is a filter very similar to \verb$itk::CastImageFilter$ excepted that it
+take cares to clip the too small or to great values, instead of simply casting them in the output
+pixel type.
+
+As \verb$itk::CastImageFilter$, it is able to process zero pixel if the output and the input are of the
+same type and the filter can be run in place.
+
 \subsection{RegionFromReferenceImageFilter}
@@
 
+\subsection{FFTConvolutionImageFilterBase}
+
+\verb$itk::FFTConvolutionImageFilter$ is the base class to use for all the FFT based convolution transforms. It
+contains several methods designed to make easier the implementation of the FFT based convolution filters. There
+are 
+\begin{itemize}
+ \item several kind of methods to prepare the input image and kernel;
+ \item a method to register the progress of some additional filters;
+ \item an easy access to the padded region and the oddity of the size on the x axis;
+ \item several kind of methods to produce the output.
+\end{itemize}
+
+It is implemented as a subclass of \verb$itk::FFTPadImageFilter$ to expose the padding options and avoid
+code duplication.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[scale=0.35]{inheritance}
+\caption{New and modified classes inheritances.\label{fig:conv-pipeline}}
+\end{figure}
+
 \subsection{FFTConvolutionImageFilter}
@@
 
+It is based on \verb$itk::FFTConvolutionImageFilter$.
+
 It sets the \verb$ReleaseDataFlag$ properly to keep the memory usage as low as possible, and propagates the
 user specified number of threads.
-
-It also exposes the \verb$GreatestPrimeFactor$ accessors from \verb$itk::FFTPadImageFilter$ to let
-the user choose between the extra performance added by the padding and the extra memory cost, and
-\verb$PadMethod$ from the same filter to let the user choose the padding method: Zero flux Neumann
-is used by default.
 
@@
 
-Unlike most of my previous contributions, this work has {\em not} been supported by
-the french National Institute for Agricultural Research (INRA) -- my employer.
-I'd like to thank the INRA anyway, and more specifically the MIMA2 facility, who
-provided the computer ressources used for this work.
-
 I'm grateful to Dan White for its corrections and useful comments on this article.
@@
  \item a more memory efficient implementation of the convolution filter using the Overlap-Save algorithm;
- \item allow the user to create a plan with \verb$FFTW_PATIENT$, \verb$FFTW_MEASURE$, or
-       \verb$FFTW_EXHAUSTIVE$, so that FFTW can create better plan with \verb$FFTW_ESTIMATE$ in a
-       later run;
  \item integrate FFTW's wisdom to improve FFT performance in ITK.
